---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
```

# ipumsr #
[![Project Status:Active](http://www.repostatus.org/badges/latest/active.svg)](http://www.repostatus.org/#active)
[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/ripums)](http://cran.r-project.org/web/packages/ripums) 
[![Travis-CI Build Status](https://travis-ci.org/mnpopcenter/ripums.svg?branch=master)](https://travis-ci.org/mnpopcenter/ripums)
[![AppVeyor Build Status](https://ci.appveyor.com/api/projects/status/github/mnpopcenter/ripums?branch=master&svg=true)](https://ci.appveyor.com/project/mnpopcenter/ripums)
[![Coverage Status](https://codecov.io/gh/mnpopcenter/ripums/master.svg)](https://codecov.io/github/mnpopcenter/ripums?branch=master)

The ripums package helps import IPUMS extracts from the [IPUMS website](https://www.ipums.org) 
into R.

The ripums package can be installed by running the following command:
```{r, eval=FALSE}
install.packages("ripums")
```

Or, you can install the development version by:
```{r, eval=FALSE}
if (!require(devtools)) install.packages("devtools")

devtools::install_github("mnpopcenter/ripums")
```

## Vignettes ##
There are several vignettes included in the package:

- **ripums** - Provides general information about using the ripums package
- **value-labels** - Provides guidance for using the value labels provided by IPUMS
- **ipums-geograph** - Provides guidance for using R as GIS tool with IPUMS data
- **ipums-cps** - An example of using CPS data with the ripums package
- **ipums-nghis** - An example of using NHGIS data with the ripums package

You can access them with the `vignette()` command (eg `vignette("value-labels")`).

If you are installing from github and want the vignettes, you'll need to run the following 
commands first:
```{r, eval = FALSE}
devtools::install_github("mnpopcenter/ripums/ripumsexamples")
devtools::install_github("mnpopcenter/ripums", build_vignettes = TRUE)
```

## Examples ##
```{r}
suppressPackageStartupMessages({
  library(ripums)
  library(haven)
  library(ggplot2) # ggplot2 version > 2.2.1 (development version as of 8/15/2017)
  library(dplyr)
  library(sf)
})
```

### CPS - Hierarchical Data ###
Relies on user downloading the .xml DDI file and the .dat/.dat.gz file (doesn't need to be unzipped).
```{r}
# Use example file included with package:
cps_hier_file <- ripums_example("cps_00010.xml")
ddi <- read_ipums_ddi(cps_hier_file)
data <- read_ipums_micro(ddi)

# Variable description for the month variable
cat(ipums_var_desc(ddi, MONTH))

# Hierarachical data loaded as a data frame


# Value labels loaded as haven::labelled class
# Convert to factors with `as_factor`
table(as_factor(data$MONTH, levels = "both"))

# Can also load as a list by rectype
data <- read_ipums_micro_list(ddi, verbose = FALSE)
# Household data
data$HOUSEHOLD
# Person data
data$PERSON
```


### CPS - Rectangular Data ###
Relies on user downloading the .xml DDI file and the .dat/.dat.gz file (doesn't need to be unzipped).
```{r, eval = FALSE}
# Use example file included with package
cps_rect_file <- ripums_example("cps_00006.xml")
data <- read_ipums_micro(cps_rect_file, verbose = FALSE)

# While working interactively, can get convenient display of variable information
# in RStudio's viewer
ipums_view(data)
```

### NHGIS ###
Relies on user downloading the csv file (with or without header row) and shape files (doesn't need to be unzipped).

Note that to save space when including this data on CRAN, the shape file has
been reduced to 1% of the points in the polygon of the PMSA. The original
shape file can be found in the `ripumsexamples` package.
```{r nhgisplot}
data <- read_nhgis_sf(
  ripums_example("nhgis0008_csv.zip"),
  shape_file = ripums_example("nhgis0008_shape_small.zip"),
  verbose = FALSE
)

ipums_var_info(data, starts_with("D6Z"))

data <- data %>%
  mutate(
    pct_before_1950 = (D6Z007 + D6Z008) / 
           (D6Z001 + D6Z002 + D6Z003 + D6Z004 + D6Z005 + D6Z006 + D6Z007 + D6Z008)
  )

# Note the function `geom_sf()` is currently only in the development version, 
# so you may need to update ggplot2 to run using 
#   devtools::install_github("tidyverse/ggplot2")
if ("geom_sf" %in% getNamespaceExports("ggplot2")) {
  ggplot(data = data) + 
    geom_sf(aes(fill = pct_before_1950)) + 
    labs(
      title = "Percent of homes built before 1950", 
      subtitle = "By Primary Metropolitan Statistical Area in 1990 Census", 
      caption = "Simplified PMSA boundaries (1% of polygon points retained)"
    )
}
```

### Terrapop ###
There is experimental support for for loading terrapop data, but examples are too large to include in the package.
```{r, eval=FALSE}
# Raster data
data <- ripums:::read_terra_raster(
  "2552_bundle.zip",
  "CROPLAND2000ZM2013.tiff",
  verbose = FALSE
)

# Area data
data <- ripums:::read_terra_area(
  "2553_bundle.zip",
  verbose = FALSE
)

# Microdata
data <- ripums:::read_terra_micro(
  "2554_bundle.zip",
  verbose = FALSE
)

```
