---
title: "Interacting with the IPUMS USA microdata extract API"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ipums-api}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Warning: New features currently in development

The IPUMS USA microdata extract API is still in testing and is not publicly
available. A call for beta testers will go out to all IPUMS USA users in late
2021 or early 2022. Moreover, the ipumsr functions for interacting with the API
are only included in the development version of the package on GitHub, not in 
the version on CRAN. If you are a beta tester, you can download the development 
version of ipumsr, which includes API functions, using:

```{r}
if (!require(remotes)) install.packages("remotes")
remotes::install_github("mnpopcenter/ipumsr", build_vignettes = TRUE)
```


# Overview

The IPUMS USA microdata extract API allows registered IPUMS USA users to define 
extracts, submit extract requests, and download extract files without visiting 
the IPUMS website. ipumsr includes functions that make it easy for R users to 
interact with the extract API from their R session.

```{r setup, eval=TRUE, message=FALSE}
library(ipumsr)
library(dplyr) # not necessary to use API functions, but used in some examples
library(purrr) # not necessary to use API functions, but used in some examples
```

# Setting up your API key

If you don’t have an IPUMS USA account, 
[register for access](https://uma.pop.umn.edu/usa/user/new?return_url=https%3A%2F%2Fusa.ipums.org%2Fusa-action%2Fmenuor).

Once you're registered, you'll need to 
[create an API key](https://account.ipums.org/api_keys).

Once you've created an API key, you can choose to supply it as a function 
argument whenever interacting with the API, or you can set the value of the 
`IPUMS_API_KEY` environment variable to your key. One convenient way to set this 
environment variable is to add code like the following to a file named 
".Renviron" in your project directory or user home directory:

```
IPUMS_API_KEY = "paste-your-key-here"
```

If you use this approach with a project that uses Git for version control, be 
sure to add the ".Renviron" file to ".gitignore" so that your API key is not 
shared when someone clones your repository.


# Defining your extract

The `define_extract()` function returns an object of class "ipums_extract" which
can then be submitted using the `submit_extract()` function.

```{r eval=TRUE}
extract_definition <- define_extract(
  collection = "usa",
  description = "Extract for API vignette",
  samples = c("us2018a","us2019a"),
  variables = c("AGE","SEX","RACE","STATEFIP"),
  data_format = "fixed_width",
  data_structure = "rectangular",
  rectangular_on = "P"
)
```

Note that samples are specified using special sample ID codes, which can be 
browsed [here on the IPUMS USA website](https://usa.ipums.org/usa-action/samples/sample_ids).

The `data_format`, `data_structure`, and `rectangular_on` arguments have default 
values which match the values specified in the example above, and you can omit 
the argument names as long as you maintain the proper order, so you could define 
the same extract with:

```{r}
extract_definition <- define_extract(
  "usa",
  "Extract for API vignette",
  c("us2018a","us2019a"),
  c("AGE","SEX","RACE","STATEFIP")
)
```


# Submitting your extract

To submit your extract, use:

```{r}
submit_extract(extract_definition)
```


However, like the `define_extract()` function, the `submit_extract()` function
returns an "ipums_extract" object, and the returned object has been updated to
include the extract number, so it can be convenient to save that return object
by assigning a name to it, like this:

```{r}
submitted_extract <- submit_extract(extract_definition)
```

That way, you can use the `submitted_extract` object as input to check the
extract's status, as shown in the next section. 

You can also get the extract number with:

```{r}
submitted_extract$number
```

If you forget to capture the return value of `submit_extract()`, you can get 
all information on your most recent extract (including the extract number) with:

```{r}
submitted_extract <- get_last_extract_info("usa")
```

`get_last_extract_info()` is just a convenience wrapper around 
`get_recent_extracts_info_list()`, described [below](#recent).


# Checking the status of your extract

To retrieve the latest status of an extract, you can use the
`get_extract_info()` function. `get_extract_info()` returns an "ipums_extract"
object with the "status" element updated to reflect the latest status of the
extract, and the "download_links" element updated to include links to any
extract files that are available for download. The "status" of a submitted
extract is one of "queued", "started", "produced", "canceled", "failed", or
"completed". Only "completed" extracts can be downloaded, but "completed"
extracts older than 72 hours may not be available for download, since extract
files are removed after that time (see discussion of `is_extract_ready()`
function below).

If you assigned a name to the return value of `submit_extract()`, as shown
above, you could get updated information on the extract, returned as an
"ipums_extract" object, with:

```{r}
submitted_extract <- get_extract_info(submitted_extract)
```

To print the latest status, you can use:

```{r}
submitted_extract$status
```

If you forget to capture the return value of `submit_extract()`, you can get 
all information on your most recent extract (including the extract number) with:

```{r}
submitted_extract <- get_last_extract_info("usa")
```

`get_last_extract_info()` is just a convenience wrapper around 
`get_recent_extracts_info_list()`, described [below](#recent).

If you don't have an "ipums_extract" object in your environment that describes
the extract you're interested in, and you don't want the most recent extract,
you can also query the latest status of an extract by supplying the name of the
IPUMS data collection and extract number of the extract, in one of two formats:

```{r}
submitted_extract <- get_extract_info("usa:33")
```

or

```{r}
submitted_extract <- get_extract_info(c("usa", "33"))
```

Note that in the first format, there are no spaces before or after the colon, 
and that in both formats, there is no need to zero-pad the extract number -- in 
other words, use "33", not "00033".

If you want R to periodically check the status of your extract, and only return 
an updated "ipums_extract" object once the extract is ready to download, you 
can use `wait_for_extract()`, as shown below:

```{r}
downloadable_extract <- wait_for_extract(submitted_extract)
```

`wait_for_extract()` also accepts the same `"collection:number"` and 
`c("collection", "number")` specifications shown above:

```{r}
downloadable_extract <- wait_for_extract("usa:33")
```

or

```{r}
downloadable_extract <- wait_for_extract(c("usa", "33"))
```

For large extracts that take a long time to produce, or when the IPUMS servers
are busy, you may not want to use `wait_for_extract()`, as it will tie up your R
session until the extract is ready to download. Alternatively, you can use the
`timeout_seconds` argument to set the maximum number of seconds you want the
function to wait. By default, that argument is set to 10,800 seconds (3 hours).

One additional way to check whether your extract is ready to download is using 
the `is_extract_ready()` function. This function accepts either an 
"ipums_extract" object or a `"collection:number"` or `c("collection", "number")`
specification, and returns a single `TRUE` or `FALSE` value indicating whether 
the extract is ready to be downloaded.

```{r}
is_extract_ready(submitted_extract)
is_extract_ready("usa:33")
is_extract_ready(c("usa", "33"))
```


# Downloading your extract

Once your extract is ready to download, use the `download_extract()` function to 
download the data and DDI codebook files to your computer. The 
`download_extract()` function returns the path to the DDI codebook file, which 
can be used to read in the downloaded data with ipumsr functions. By default, 
the function will download files into your current working directory, but 
alternative locations can be specified with the `download_dir` argument.

```{r}
ddi_path <- download_extract(submitted_extract)

# Assuming you chose either "fixed-width" or "csv" for your extract's "data_format"
ddi <- read_ipums_ddi(ddi_path)
data <- read_ipums_micro(ddi)
```


Or, using a `"collection:number"` or `c("collection", "number")` specification:

```{r}
ddi_path <- download_extract("usa:33")
ddi_path <- download_extract(c("usa", "33"))
```


# Sharing an extract definition

One exciting feature enabled by the IPUMS USA data extract API is the ability to 
share a standardized extract definition with other IPUMS users so that they can 
create an identical extract for themselves. ipumsr facilitates this by offering 
the functions `save_extract_as_json()` and `define_extract_from_json()` to 
write extract definitions to and read extract definitions from a standardized 
JSON-formatted file.

To write the definition of your USA extract number 33 to a JSON-formatted file 
that can be shared with other users, you could use:

```{r}
usa_extract_33 <- get_extract_info("usa:33")
save_extract_as_json(usa_extract_33, file = "usa_extract_33.json")
```

Then, assuming that file is saved in your current working directory, you or
another user could use that JSON file to create a duplicate "ipums_extract"
object with the same definition, and submit it, using:

```{r}
clone_of_usa_extract_33 <- define_extract_from_json("usa_extract_33.json", "usa")
submitted_extract <- submit_extract(clone_of_usa_extract_33)
```


# Revising a previous extract

ipumsr also includes a convenience function for revising a previous extract 
definition, facilitating a "revise and resubmit" workflow. Here's how you would 
pull down the definition of USA extract number 33 and add a sample and a 
variable to it:

```{r}
old_extract <- get_extract_info("usa:33")
new_extract <- revise_extract(
  old_extract,
  samples_to_add = "us2018a",
  vars_to_add = "RELATE"
)
```

The `revise_extract()` function returns an "ipums_extract" object that has been 
modified as requested and has been reset to an unsubmitted state, by stripping 
the extract number, status, and download links from the original extract. The 
revised extract can then be submitted with:

```{r}
newly_submitted_extract <- submit_extract(new_extract)
```


# Getting info on multiple recent extracts {#recent}

You can query the API for the details and status of up to ten recent extracts 
using the functions `get_recent_extracts_info_list()` and 
`get_recent_extracts_info_tbl()`. The `_list` version of the function returns a 
list of "ipums_extract" objects, whereas the `_tbl` version returns a tibble 
(enhanced "data.frame") in which each row contains information on one extract. 

The list representation is useful if you want to be able to operate on elements
as "ipums_extract" objects. For instance, to retrieve your most
second-most-recent extract and revise it for resubmission, you could use:

```{r}
second_most_recent_extract <- get_recent_extracts_info_list("usa")[[2]]
revised_extract <- revise_extract(
  second_most_recent_extract, 
  samples_to_add = "us2010a"
)
```

Or to download all recent extracts that are ready to download, using 
`purrr::keep()` and `purrr::map_chr()`:

```{r}
ddi_paths <- get_recent_extracts_info_list("usa") %>% 
  keep(is_extract_ready) %>% 
  map_chr(download_extract)
```

The tibble representation is useful if you want to use functions for 
manipulating data.frames to find recent extracts matching particular criteria.

```{r}
recent_usa_extracts_tbl <- get_recent_extracts_info_tbl("usa")
```

For example, to find extracts with descriptions including the word "occupation", 
you could use:

```{r}
recent_usa_extracts %>%  
  filter(grepl("occupation", description))
```

Filtering on properties such as "samples" or "variables" is a little more
complex, because these are stored in list columns, but it is possible. For
example, to find extracts including the variable "AGE", you could use
`purrr::map_lgl()` like this:

```{r}
recent_usa_extracts %>% 
  filter(map_lgl(variables, ~"AGE" %in% .x))
```

To convert between these two representations, ipumsr provides the functions 
`extract_list_to_tbl()` and `extract_tbl_to_list()`, such that the following is 
`TRUE`:

```{r}
identical(
  extract_list_to_tbl(get_recent_extracts_info_list("usa")),
  get_recent_extracts_info_tbl("usa")
)
```


# Putting it all together, with pipes

The return values of the functions to interact with the API are configured in 
such a way that you can define, submit, wait for, download, and read in your 
extract all in one piped expression:

```{r}
data <- 
  define_extract(
    "usa",
    "Extract for API vignette",
    c("us2018a","us2019a"),
    c("AGE","SEX","RACE","STATEFIP")
  ) %>% 
    submit_extract() %>% 
    wait_for_extract() %>% 
    download_extract() %>% 
    read_ipums_micro()
```
