---
title: "Interacting with the IPUMS USA microdata extract API"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ipums-api}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Warning: New features currently in development

The IPUMS microdata extract API is still in beta testing, and the interface 
for interacting with the API with ipumsr could still change in response to 
tester feedback. If you are interested in becoming a beta tester, email 
ipums+api@umn.edu.

To install the latest version of ipumsr from CRAN, use:

```{r}
install.packages("ipumsr")
```

# Overview

The IPUMS microdata extract API allows registered IPUMS USA and CPS users to
define extracts, submit extract requests, and download extract files without
visiting the IPUMS website. ipumsr includes functions that help R users interact
with the extract API from their R session.

```{r setup, eval=TRUE, message=FALSE}
library(ipumsr)
library(dplyr) # not necessary to use API functions, but used in some examples
library(purrr) # not necessary to use API functions, but used in some examples
```

# Setting up your API key

If you don’t have an IPUMS USA account, 
[register for access](https://uma.pop.umn.edu/usa/user/new?return_url=https%3A%2F%2Fusa.ipums.org%2Fusa-action%2Fmenuor).

Once you're registered, you'll need to 
[create an API key](https://account.ipums.org/api_keys).

Once you've created an API key, you can choose to supply it as a function 
argument whenever interacting with the API, or you can set the value of the 
`IPUMS_API_KEY` environment variable to your key. The example code in this 
vignette assumes you have assigned your key to this environment variable.

To set the value of the `IPUMS_API_KEY` environment variable for your current 
session, you can use:

```{r}
set_ipums_api_key("paste-your-key-here")
```

To set your API key *and* save it for use in future sessions, use the same 
function, but with `save` set to `TRUE`:

```{r}
set_ipums_api_key("paste-your-key-here", save = TRUE)
```


This will add your API key to a file named ".Renviron" in your user home 
directory, so that the value of the `IPUMS_API_KEY` environment variable is 
set when R starts up.


# Defining your extract

Each IPUMS data collection with API support has its own function for 
defining an extract. These functions have names of the form 
`define_extract_<collection>()`. Thus, to define an IPUMS USA extract, you use 
`define_extract_usa()`, and to define an IPUMS CPS extract, you use 
`define_extract_cps()`. 

All `define_extract_()` functions return an `ipums_extract` object which can
then be submitted using the `submit_extract()` function. 

```{r eval=TRUE}
usa_extract_definition <- define_extract_usa(
  description = "USA extract for API vignette",
  samples = c("us2018a","us2019a"),
  variables = c("AGE","SEX","RACE","STATEFIP")
)

cps_extract_definition <- define_extract_cps(
  description = "CPS extract for API vignette",
  samples = c("cps1976_01s", "cps1976_02b"),
  variables = c("YEAR", "MISH", "CPSIDP", "AGE", "SEX", "RACE", "UH_SEX_B1")
)
```

For more details on the `ipums_extract` class, view the documentation page with 
`?ipums_extract-class`.

Note that samples are specified using special sample ID codes, which can be 
browsed [here on the IPUMS USA website](https://usa.ipums.org/usa-action/samples/sample_ids),
or [here for IPUMS CPS](https://cps.ipums.org/cps-action/samples/sample_ids).


# Submitting your extract

To submit your extract, use:

```{r}
submit_extract(usa_extract_definition)
```


However, like the `define_extract_()` functions, the `submit_extract()` function
returns an `ipums_extract` object, and the returned object has been updated to
include the extract number, so it can be useful to save that return object
by assigning a name to it, like this:

```{r}
submitted_usa_extract <- submit_extract(usa_extract_definition)
```

That way, you can use the `submitted_usa_extract` object as input to check the
extract's status, as shown in the next section, or to reference the extract 
number:

```{r}
submitted_usa_extract$number
```


# Checking the status of your extract

To retrieve the latest status of an extract, you can use the
`get_extract_info()` function. `get_extract_info()` returns an `ipums_extract`
object with the "status" element updated to reflect the latest status of the
extract, and the "download_links" element updated to include links to any
extract files that are available for download. 

The "status" of a submitted extract is one of "queued", "started", "produced",
"canceled", "failed", or "completed". Only "completed" extracts can be
downloaded, but "completed" extracts older than 72 hours may not be available
for download, since extract files are removed after that time (see discussion of
the `is_extract_ready()` function below).

If you assigned a name to the return value of `submit_extract()`, as shown
above, you could get updated information on the extract, returned as an
`ipums_extract` object, with:

```{r}
submitted_usa_extract <- get_extract_info(submitted_usa_extract)
```

To print the latest status, you can use:

```{r}
submitted_extract$status
```

If you forget to capture the return value of `submit_extract()`, you can pull 
down an `ipums_extract` object containing all the information on your most 
recent extract for a given data collection with:

```{r}
submitted_usa_extract <- get_last_extract_info("usa")
```

`get_last_extract_info()` is just a convenience wrapper around 
`get_recent_extracts_info_list()`, described [below](#recent).

If you don't have an `ipums_extract` object in your environment that describes
the extract you're interested in, and you don't want the most recent extract,
you can also query the latest status of an extract by supplying the name of the
IPUMS data collection and extract number of the extract, in one of two formats. 
Here's how you'd get the latest information on IPUMS CPS extract number 33:

```{r}
cps_extract_33 <- get_extract_info("cps:33")
```

or

```{r}
cps_extract_33 <- get_extract_info(c("cps", "33"))
```

Note that in the first format, there are no spaces before or after the colon, 
and that in both formats, there is no need to zero-pad the extract number -- in 
other words, use "33", not "00033".

If you want R to periodically check the status of your extract, and only return 
an updated `ipums_extract` object once the extract is ready to download, you 
can use `wait_for_extract()`, as shown below:

```{r}
downloadable_cps_extract <- wait_for_extract(cps_extract_33)
```

`wait_for_extract()` also accepts the same `"collection:number"` and 
`c("collection", "number")` specifications shown above:

```{r}
downloadable_cps_extract <- wait_for_extract("cps:33")
```

or

```{r}
downloadable_cps_extract <- wait_for_extract(c("cps", "33"))
```

For large extracts that take a long time to produce, or when the IPUMS servers
are busy, you may not want to use `wait_for_extract()`, as it will tie up your R
session until the extract is ready to download.

`wait_for_extract()` will tie up your R session until the extract is ready to
download, so it might not be the best option for large extracts that take a
long time to produce. However, `wait_for_extract()` does offer a
`timeout_seconds` argument to set the maximum number of seconds you want the
function to wait. By default, that argument is set to 10,800 seconds (3 hours).

An alternative way to check whether your extract is ready to download is using 
the `is_extract_ready()` function. This function accepts either an 
`ipums_extract` object or a `"collection:number"` or `c("collection", "number")`
specification, and returns a single `TRUE` or `FALSE` value indicating whether 
the extract is ready to be downloaded.

```{r}
is_extract_ready(cps_extract_33)
is_extract_ready("cps:33")
is_extract_ready(c("cps", "33"))
```

As noted above, only extracts with status "completed" can be ready to download, 
but not all "completed" extracts are ready to download, because extract files 
are removed from IPUMS servers after 72 hours. The `is_extract_ready()` 
function checks whether an extract can currently be downloaded by looking at the 
"download_links" element of the extract object returned by the API.

Note that the API has a limit of 60 requests with the same API key per minute, 
so you wouldn't want to write a loop that repeatedly uses `is_extract_ready()` 
to check your extract status.


# Downloading your extract

Once your extract is ready to download, use the `download_extract()` function to 
download the data and DDI codebook files to your computer. The 
`download_extract()` function returns the path to the DDI codebook file, which 
can be used to read in the downloaded data with ipumsr functions. By default, 
the function will download files into your current working directory, but 
alternative locations can be specified with the `download_dir` argument.

```{r}
ddi_path <- download_extract(submitted_usa_extract)

ddi <- read_ipums_ddi(ddi_path)
data <- read_ipums_micro(ddi)
```


Or, using a `"collection:number"` or `c("collection", "number")` specification:

```{r}
ddi_path <- download_extract("cps:33")
ddi_path <- download_extract(c("cps", "33"))
```


# Sharing an extract definition

One exciting feature enabled by the IPUMS microdata extract API is the ability to 
share a standardized extract definition with other IPUMS users so that they can 
create an identical extract for themselves. ipumsr facilitates this by offering 
the functions `save_extract_as_json()` and `define_extract_from_json()` to 
write extract definitions to and read extract definitions from a standardized 
JSON-formatted file.

To write the definition of your CPS extract number 33 to a JSON-formatted file 
that can be shared with other users, you could use:

```{r}
cps_extract_33 <- get_extract_info("cps:33")
save_extract_as_json(cps_extract_33, file = "cps_extract_33.json")
```

Then, you or another user could use that JSON file to create a duplicate
`ipums_extract` object with the same definition, and submit it, using:

```{r}
clone_of_cps_extract_33 <- define_extract_from_json("cps_extract_33.json")
submitted_cps_extract <- submit_extract(clone_of_cps_extract_33)
```

Note that the code in the previous chunk assumes that the file is saved in the
current working directory. If it's saved somewhere else, replace 
`"cps_extract_33.json"` with the full path to the file.


# Revising a previous extract

ipumsr also includes convenience functions for revising a previous extract 
definition, facilitating a "revise and resubmit" workflow. Here's how you would 
pull down the definition of USA extract number 33 and add a sample and a 
variable to it:

```{r}
old_extract <- get_extract_info("usa:33")
new_extract <- add_to_extract(old_extract, samples = "us2020a", vars = "RELATE")
```

The `add_to_extract()` function returns an `ipums_extract` object that has been 
modified as requested and has been reset to an unsubmitted state, by stripping 
the extract number, status, and download links from the original extract. The 
revised extract can then be submitted with:

```{r}
newly_submitted_extract <- submit_extract(new_extract)
```

To remove values from an extract, use `remove_from_extract()`:

```{r}
newer_extract <- remove_from_extract(new_extract, samples = "us2020a")
```



# Getting info on multiple recent extracts {#recent}

You can query the API for the details and status of recent extracts (the ten most 
recent, by default) using the functions `get_recent_extracts_info_list()` and 
`get_recent_extracts_info_tbl()`. The `_list` version of the function returns a 
list of `ipums_extract` objects, whereas the `_tbl` version returns a tibble 
(enhanced "data.frame") in which each row contains information on one extract. 

The list representation is useful if you want to be able to operate on elements
as `ipums_extract` objects. For instance, to retrieve your most
second-most-recent extract and revise it for resubmission, you could use:

```{r}
second_most_recent_extract <- get_recent_extracts_info_list("usa")[[2]]
revised_extract <- revise_extract_micro(
  second_most_recent_extract, 
  samples_to_add = "us2010a"
)
```

Or to download all recent extracts that are ready to download, using 
`purrr::keep()` and `purrr::map_chr()`:

```{r}
ddi_paths <- get_recent_extracts_info_list("usa") %>% 
  keep(is_extract_ready) %>% 
  map_chr(download_extract)
```

The tibble representation is useful if you want to use functions for 
manipulating data.frames to find recent extracts matching particular criteria.

```{r}
recent_usa_extracts_tbl <- get_recent_extracts_info_tbl("usa")
```

For example, to find extracts with descriptions including the word "occupation", 
you could use:

```{r}
recent_usa_extracts %>%  
  filter(grepl("occupation", description))
```

Filtering on properties such as "samples" or "variables" is a little more
complex, because these are stored in list columns, but it is possible. For
example, to find extracts including the variable "AGE", you could use
`purrr::map_lgl()` like this:

```{r}
recent_usa_extracts %>% 
  filter(map_lgl(variables, ~"AGE" %in% .x))
```

To convert between these two representations, ipumsr provides the functions 
`extract_list_to_tbl()` and `extract_tbl_to_list()`, such that the following is 
`TRUE`:

```{r}
identical(
  extract_list_to_tbl(get_recent_extracts_info_list("usa")),
  get_recent_extracts_info_tbl("usa")
)
```


# Putting it all together, with pipes

The return values of the functions to interact with the API are configured in 
such a way that you can define, submit, wait for, download, and read in your 
extract all in one piped expression:

```{r}
data <- 
  define_extract_usa(
    "USA extract for API vignette",
    c("us2018a","us2019a"),
    c("AGE","SEX","RACE","STATEFIP")
  ) %>% 
    submit_extract() %>% 
    wait_for_extract() %>% 
    download_extract() %>% 
    read_ipums_micro()
```
