---
title: "Geographic Data in `ripums`"
author: "Minnesota Population Center"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ipums-geography}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

# Don't give instructions about downloading data for yourself for geography data
# because some of the files (like the small shape file) aren't available.
ripumsexamples_file <- system.file("extdata", "nhgis0010_csv.zip", package = "ripumsexamples")

if (!file.exists(ripumsexamples_file)) {
  message(paste0(
    "Could not find data and so could not run vignette.\n\n",
    "The data is available on github - you can install it using the following ",
    "commands: \n",
    "  if (!require(devtools)) install.packages('devtools')\n",
    "  devtools::install_github('mnpopcenter/ripums/ripumsexamples')\n",
    "After installation, the data should be available for this vignette.\n\n"
  ))
  knitr::opts_chunk$set(eval = FALSE)
}

if (!suppressPackageStartupMessages(require(sf))) {
  message("Could not find sf package and so could not run vignette.")
  knitr::opts_chunk$set(eval = FALSE)
}
```
IPUMS geographic data takes several different forms:

- Microdata projects like 
[IPUMS DHS](https://www.idhsdata.org/idhs/gis.shtml), 
[IPUMS International](https://international.ipums.org/international/geography_gis.shtml) 
and [IPUMS USA](https://usa.ipums.org/usa/volii/tgeotools.shtml) provide shape 
files on dedicated geography pages that are separate from the microdata extract system.
- [IPUMS NHGIS](https://nhgis.org) provides access to the shape files relevant to your
current extract as part of the system to create the data extract.
- [IPUMS Terra](https://www.terrapop.org/) provides shape files along with the extract
and allows for aggregating microdata and raster data into the boundaries. 

(Note: The ripums package does not yet support IPUMS Terra data. We hope to add
support for IPUMS Terra soon.)

This vignette offers an introduction to the types of geographies available through IPUMS, then uses example data to walk through two typical workflows for using boundary data with (1) IPUMS DHS, IPUMS International or IPUMS USA, and (2) IPUMS NHGIS. The vignette focuses on the specifics of working with IPUMS geographic data, and is by no means an overview of working with geographic data in R. In the examples, I show how to make static graphs using ggplot2, but there are many other R packages to help you
do all sorts of things with geographic data, including making interactive maps and performing geospatial analysis. [At the end of
this vignette](#resources), there are some resources for learning more.

## Introduction to IPUMS Geographies
IPUMS geographic data can be classified along two dimensions. 

### Harmonized vs. Non-Harmonized
Changing boundaries complicate longitudinal analyses of geographic units. For many geographies
IPUMS provides boundary files that have been made consistent by combining
geographies that share area for different time periods (see the project specific
geography webpages for more details).
IPUMS geography pages refer to these as "harmonized", "integrated" or "consistent". 
Boundaries that have not been made consistent over time are often referred to as "year-specific," or may simply be labelled with the year to which they pertain (eg "2010 PUMAs" for IPUMS USA).

### Level
The specificity of the geography (eg states and
counties in the United States). IPUMS often provides multiple levels of 
data, sometimes overlapping (as the case of counties within states) but sometimes
not (eg zip codes in the US may not fit within county boundaries).

## A Note on R's Geospatial Packages: `sf` vs `sp`
The `ripums` package allows for loading geospatial data in two formats (sf for
Simple Features and sp for Spatial). The `sf` package is relatively new, and so 
does not have as widespread support as the `sp` package. However, (in my opinion)
it does allow for easier analysis, and so may be a better place to start if you
have not used GIS data in R before.

All examples in this vignette will use the sf package, but sp's functionality
is generally similar. The only place where the code is noticeably different
is for making plots. See the [resources at the end of this vignette](#resources) for
some instructions on plotting with the sp package.

## Workflow for IPUMS DHS, IPUMS International, and IPUMS USA
Every analysis will be slightly different, but in general the workflow for integrating geographic data and microdata from IPUMS DHS, IPUMS International or IPUMS USA can be divided into eight steps:

1. Determine what geographic data you want to use (and is available)
2. Download microdata and shape files
3. Load microdata and shape files into R
4. Manipulate microdata as needed
5. Summarize microdata by geographic level
6. Combine multiple shape files (if necessary)
7. Merge microdata and boundary data
8. Analyze the data!

Below, I will walk through these steps for an example that maps geographic variation in the use of solid fuel
for cooking in households with children aged 0-5 in Colombia, 
Ecuador and Peru. 

The data for this example is included in the package `ripumsexamples`,
which is available on github. If you want to run the code from this 
vignette you will need to install it by running the following commands:
```{r, eval = FALSE}
if (!require(devtools)) install.packages(devtools)
devtools::install_github("mnpopcenter/ripums/ripumsexamples")
```

```{r}
# This function helps find the data from the ripumsexamples package for us
ex_file <- function(x) {
  system.file("extdata", x, package = "ripumsexamples")
}
```

Before you work on the data for yourself, I recommend reading
the specifics about the project's geographic data on the IPUMS website
([IPUMS DHS](https://www.idhsdata.org/idhs/gis.shtml) 
[IPUMS International](https://international.ipums.org/international/geography_gis.shtml) 
and [IPUMS USA](https://usa.ipums.org/usa/volii/tgeotools.shtml)).
Each project has its own idiosyncrasies that you'll want to understand
before diving in.


### 1) Determine what geographic data you want to use (and is available) 
From the data extract engine I see that the variable FUELCOOK is available
in three censuses from Colombia (1985, 1993 and 2005), and one each 
from Ecuador (2010) and Peru (2007).

The decision to use harmonized or year-specific geographies depends on whether you want to analyze change over time. If I want to
be able to compare across time in Colombia, I will need to use the harmonized
version of the Colombian geography. For Ecuador and Peru, on the other hand, FUELCOOK is only available in one census year, so there is no clear benefit to using harmonized geographies, and in fact I may lose geographic detail by doing so.

Similarly, the level of geography you choose depends on your research question. For this example I will use level 1 geography.

### 2) Download microdata and shape files
Once I know which geographic data I want, I need to add those geographic variables to my microdata extract *and* download the corresponding shape files from the project specific geography page.

In the IPUMS International microdata extract system, I added the variable GEOLEV1 (which contains harmonized first-level geography) for Colombia, and the year-specific variables GEO1_EC2010 and GEO1_PE2007 for Ecuador and Peru, respectively.

To download shape files on the [IPUMS International geography page](https://international.ipums.org/international/geography_gis.shtml), I can click on the "GIS BOUNDARY FILES DOWNLOAD" link. For Colombia, I'll want to look on the "Spatially harmonized first-level geography" page, whereas for Ecuador and Peru, I will browse the "Year-specific first-level geography" page. On these pages, I can download the 1964-2005 harmonized shape files for Colombia, the 2010 shape files for Ecuador, and the 2007 shape files for Peru.

### 3) Load microdata and shape files into R
The `read_ipums_micro()` function loads the data and the `read_ipums_sf()` 
(or `read_ipums_sp()`) function loads the shape files. 

```{r}
suppressPackageStartupMessages({
  library(ripums)
  library(sf)
  library(dplyr)
  library(ggplot2)
})

# Load data
ipumsi_ddi_file <- ex_file("ipumsi_00011.xml")
ipumsi_ddi <- read_ipums_ddi(ipumsi_ddi_file)
ipumsi_data <- read_ipums_micro(ipumsi_ddi_file, verbose = FALSE)

# Load shape files
colombia_shape <- read_ipums_sf(ex_file("geo1_co1964_2005.zip"), verbose = FALSE)
ecuador_shape <- read_ipums_sf(ex_file("geo1_ec2010.zip"), verbose = FALSE)
peru_shape <- read_ipums_sf(ex_file("geo1_pe2007.zip"), verbose = FALSE)
```

### 4) Manipulate microdata as needed
Though IPUMS microdata has been cleaned and formatted to be easy to use
(as discussed in the value-labels vignette), generally there is at least
some data processing we need to do. 

In this example, I want to make the solid fuel use variable binary and
also use the value labels for COUNTRY, SAMPLE and AGE2, but zap them for 
the geography variables. 

When combining with shape files, the shape files also generally have the labels
for the geographic units, so we do not need to keep the labels on the geographic variables in the microdata. Also, because the
name of a geography may not be unique (eg there are 8 states in the US with a
Fulton County), it's better to merge on the unique numeric code.
```{r}
# Convert labelled values to factors where useful (and zap elsewhere)
# See the value-labels vignette for more on this process.
fuel_labels <- ipums_val_labels(ipumsi_data$FUELCOOK)
fuel_missing_lbls <- c(
  "NIU (not in universe)", "Multiple fuels", "Other combinations", "Other", "Unknown/missing"
)
fuel_solid_vals <- c(50:56, 61, 73, 74, 75)

ipumsi_data <- ipumsi_data %>%
  mutate_at(vars(COUNTRY, SAMPLE, AGE2), ~as_factor(lbl_clean(.))) %>%
  # We will get labels from shape file for geographic variables
  mutate_at(vars(starts_with("GEO")), zap_labels) %>% 
  mutate(
    SOLIDFUEL = FUELCOOK %>% 
      lbl_na_if(~.lbl %in% fuel_missing_lbls) %>%
      lbl_relabel(
        lbl(0, "Non-solid Fuel") ~ !.val %in% fuel_solid_vals,
        lbl(1, "Solid Fuel") ~ .val %in% fuel_solid_vals
      ) %>%
      as_factor(),
    FUELCOOK = as_factor(FUELCOOK)
  )
```


### 5) Summarize microdata by geographic level
Microdata has one observation per person. To make a map
we have to summarize across persons by geographic
unit. Since we have data at three time points for Colombia, we will summarize by geographic unit and year.

Because in this example we are using both harmonized and year-specific geographic units, the
data currently has multiple variables representing the geographic ID.
Now is a good chance to combine them into one.

Here we calculate the percent of all children who live in a household
that uses solid fuel for cooking.
```{r}
ipumsi_summary <- ipumsi_data %>% 
  mutate(GEOLEV1 = case_when(
    COUNTRY == "Colombia" ~ GEOLEV1,
    COUNTRY == "Ecuador" ~ GEO1_EC2010,
    COUNTRY == "Peru" ~ GEO1_PE2007
  )) %>%
  group_by(YEAR, COUNTRY, GEOLEV1) %>%
  summarize(pct_solid = mean(SOLIDFUEL == "Solid Fuel", na.rm = TRUE))
```

### 6) Combine multiple shape files (if necessary)
In this example we have three shape files. The command `rbind()` has been
implemented for sf (and sp) spatial objects, but before we do that, we want to
give each country's shape data the same structure. When combining non-harmonized
geographies or across different geographic levels, the variable names will not
be the same.

```{r}
# Currently each shape file has different variable names
names(colombia_shape)
names(ecuador_shape)
names(peru_shape)

# Keep CNTRY_NAME (because the year-specific geography codes are not unique across
# countries, so we need to merge using it), ADMIN_NAME (to get the name of 
# geography), and rename GEOLEVEL1, IPUM2010 and IPUM2007 to the same variable
# name (and geography, which contains the shape)
colombia_shape <- colombia_shape %>%
  select(CNTRY_NAME, ADMIN_NAME, GEOJOIN = GEOLEVEL1)
ecuador_shape <- ecuador_shape %>%
  select(CNTRY_NAME, ADMIN_NAME, GEOJOIN = IPUM2010)
peru_shape <- peru_shape %>%
  select(CNTRY_NAME, ADMIN_NAME, GEOJOIN = IPUM2007)

# Now we can rbind them together
all_shapes <- rbind(colombia_shape, ecuador_shape, peru_shape)
```

### 7) Merge microdata and boundary data
The function `ipums_shape_inner_join()` allows us to combine the data and 
shape file while smoothing out some of the complications because the data
and shape files are provided in slightly different formats (eg,
the geography ID variable is numeric in the microdata, but stored as character in the
shape file). It also gives a message when some of the data has been
dropped by the join.

```{r}
ipumsi <- ipums_shape_inner_join(
  ipumsi_summary, 
  all_shapes,
  by = c("COUNTRY" = "CNTRY_NAME", "GEOLEV1" = "GEOJOIN")
)
```

Notice that the join warned us about one observation dropping out of the shape
file. Because we chose an inner join, all observations that were not in both the
microdata and the shape data are dropped from the results. Join failures like this
can happen for a variety of reasons. It can mean that something is wrong with
your data, perhaps you merged the wrong shape file, or didn't load an entire
file. However, this can also happen when there is a geographic unit in the shape files for which there are no microdata observations, or there is a geographic unit in the microdata for which no shape file is available, perhaps because it is too small.

We can investigate this join failure using the `join_failures` command.
```{r}
join_failures(ipumsi)
```

In this case, it appears that the shape file includes Lake Titicaca for mapping
purposes, even though no one lives in the lake. We will just leave this boundary
out of our dataset.

### 8) Analyze the data!
Now the data is ready for us to use. Here's a quick example of how to
make a map with ggplot2.
```{r, fig.height = 4, fig.width = 7}
# Note the function `geom_sf()` is a very new function, so you may need to update
# ggplot2 to run.
ipumsi <- ipumsi %>%
  mutate(census_round = cut(YEAR, c(1984, 1992, 2004, 2014), c("1985", "1993", "2005-2010")))

if ("geom_sf" %in% getNamespaceExports("ggplot2")) {
  ggplot(data = ipumsi, aes(fill = pct_solid)) +
    geom_sf() + 
    facet_wrap(~census_round) + 
    scale_fill_continuous("", labels = scales::percent) + 
    labs(
      title = "Percent of Children 0-5 Who Live in a Home That Cooks Using Solid Fuel",
      subtitle = "Colombia (1985, 1993, 2005), Ecuador (2010) and Peru (2007) Census Data",
      caption = paste0("Source: ", ipums_file_info(ipumsi_ddi, "ipums_project"))
    )
}
```

## Workflow for IPUMS NHGIS
The geographic data is more tightly integrated with the census data for
the IPUMS NHGIS project, but otherwise the process for using IPUMS NHGIS data is
similar to using the microdata projects.

Again, the specifics of each analysis will be slightly different, but
here is an example showing the median age by census block in Connecticut
and Rhode Island.

### 1) Getting the data and shape files
IPUMS NHGIS has better integration of the shape files and data. The website's extract
engine allows you to filter on variables that are only available at a particular
geographic level when looking for variables and to select particular levels
after selecting a table. Also, the geographic boundary files can be selected
alongside the tables using the "GIS Boundary Files" tab. The extract engine also
provides a single place to download both the data and the boundary files.

In our example, we select table P13, "Median Age by Sex" at the census block level.
Because census block data is so large, the extract engine divides each state's 
boundaries into different files, so we also need to select that we want Connecticut
and Rhode Island.

### 2) Load data and shape files into R
For most uses you will be able to load the data and shape files at the same
time using the `read_nhgis_sf()` (or `read_nhgis_sp()`) function. Additionally,
when the geographic data is only divided because of size reasons (like 
how each census block file is for a single state), these functions will
combine the shape files for us. **Derek comment: are there situations where these functions won't combine shape files? The current phrasing suggests that if the data are split across files for a reason other than size, the functions might not automatically combine them.**

So in our example, we run the following to load the data:
```{r}
nhgis_ddi <- read_ipums_codebook(ex_file("nhgis0024_csv.zip")) 
nhgis <- read_nhgis_sf(
  data_file = ex_file("nhgis0024_csv.zip"),
  shape_file = ex_file("nhgis0024_shape_small.zip"),
  verbose = FALSE
)
```

In other situations, such as combining data from different tables or
geographic level, you may need to load the data and shape files
separately. In this case, you can use the function `read_nhgis()` to load
the data and `read_ipums_sf()` (or sp) to load the shape files. Then the
process for combining the data is similar to the process described above
for the IPUMS International data.

### 3) Analyze the data!
Now we're ready to use the data. Here is a map of the median age for
Hartford County, CT and Providence County, RI.
```{r, fig.height = 4, fig.width = 7}
# The median age is 0 for unpopulated counties, set them to NA
nhgis <- nhgis %>%
  mutate(H77001 = ifelse(H77001 == 0, NA, H77001))

# For map filter to Hartford County, CT and Providence County, RI
nhgis_subset <- nhgis %>%
  filter(COUNTY %in% c("Hartford County", "Providence County")) %>%
  mutate(place_name = paste0(COUNTY, ", ", STATE))

if ("geom_sf" %in% getNamespaceExports("ggplot2")) {
  ggplot(data = nhgis_subset, aes(fill = H77001)) +
    geom_sf(linetype = "blank") + 
    scale_fill_continuous("") + 
    facet_wrap(~place_name, scales = "free") + 
    labs(
      title = "Median Age of Population By Census Block",
      subtitle = "2010 Census",
      caption = paste0(
        "Source: ", ipums_file_info(nhgis_ddi, "ipums_project"), "\n",
        "Simplified Census Block boundaries (1% of points retained)"
      )
    )
}
```

## Resources for Geographic Analysis in R {#resources}
There are so many great resources on working with geographic data in R, 
it's hard to pick just a few. Here are a few places to start:

- For general information about the sf and sp packages, I recommend reading the 
vignettes for those packages(`vignette(package = "sf")` and 
`vignette(packge = "sp")`). 
- The [leaflet](https://rstudio.github.io/leaflet/) package is great for interactive 
web-technology based maps. 
- The CRAN website has a great tutorial called ["Introduction to visualizing 
spatial Data in R"](https://cran.r-project.org/doc/contrib/intro-spatial-rl.pdf)
that focuses on visualizing data from the sp package. 
- The [R consortium proposal for the sf 
package](https://github.com/r-spatial/sf/blob/master/PROPOSAL.md) 
provides of the history behind the sf package.
- The [CRAN Task View for Spatial Analysis](https://cran.r-project.org/web/views/Spatial.html) describes many of the packages available for spatial analysis in R.

