---
title: "`ripums` Example - CPS"
author: "Minnesota Population Center"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ipums-cps}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette is adapted from the CPS Data Training Exercise available here:
[https://pop.umn.edu/sites/pop.umn.edu/files/final_review_-_cps_spss_exercise_1_0.pdf]()

### Research Questions ##
What is the frequency of food stamp recipiency in the US? Are health and work
statuses related?

### Objectives
- Create and download an IPUMS data extract
- Decompress data file and read data into R
- Analyze the data using sample code
- Validate data analysis work using answer key

### IPUMS Variables
- PERNUM: Person number in sample unit
- FOODSTMP: Food stamp receipt
- AGE: Age
- EMPSTAT: Employment status
- HRSWORK: Hours worked last week
- HEALTH: Health status

## Download Extract from IPUMS Website
1) **Register with IPUMS** - Go to http://cps.ipums.org, click on CPS Registration and apply for access. On login screen, enter email address and password and submit it!

2) **Make an Extract**
  - Go to http://cps.ipums.org, click on CPS Registration and Apply for access
  - On login screen, enter email address and password and submit
  - Go back to homepage and go to Select Data
  - Click the Select Samples box, check the box for the 2011 ASEC sample, click the Submit sample selections box
  - Using the drop down menu or search feature, select the following variables:
    - PERNUM: Person number in sample unit
    - FOODSTMP: Food stamp receipt
    - AGE: Age
    - EMPSTAT: Employment status
    - AHRSWORK: Hours worked last week
    - HEALTH: Health status

3) **Request the Data**
  - Click the green VIEW CART button under your data cart
  - Review variable selection. Click the green Create Data Extract button
  - Review the ‘Extract Request Summary’ screen, describe your extract and click Submit Extract
  - You will get an email when the data is available to download
  - To get to the page to download the data, follow the link in the email, or follow the Download and Revise Extracts link on the homepage

4) **Download the Data**
  - Go to http://cps.ipums.org and click on Download or Revise Extracts
  - Right-click on the data link next to extract you created
  - Choose "Save Target As..." (or "Save Link As...")
  - Save into "Documents" (that should pop up as the default location)
  - Do the same thing for the DDI link next to the extract

## Getting the data into R ##
To facilitate running this vignette in R, the example code here uses
the pre-made extract included in the `ripumstest` package. If you 
would like to use the extract that you have created, change the filepaths 
noted below.

```{r}
library(ripums)
library(dplyr, warn.conflicts = FALSE)
library(haven)

cps_ddi_file <- ripums_example("cps_00011.xml")
# OR uncomment this line below (by removing '#' at the beginning)
# and change the file path to where you have saved the xml file
# cps_ddi_file <- "C:/Users/My Name/My Documents/cps_00001.xml"

cps_data_file <- ripums_example("cps_00011.dat.gz")
# OR uncomment this line below (by removing '#' at the beginning)
# and change the file path to where you have saved the xml file
# cps_ddi_file <- "C:/Users/My Name/My Documents/cps_00001.xml"

cps_data <- read_ipums_micro(cps_ddi_file, data_file = cps_data_file)
```

## Exercises ##

### Analyze the Sample – Part I Frequencies of FOODSTMP
A) On the website, find the codes page for the FOODSTMP variable
and write down the code value, and what category each code
represents. 
```{r}
# Can find on the website or from the data
print_labels(cps_data$FOODSTMP)
```
B) What is the universe for FOODSTMP in 2011 (under the
Universe tab on the website)? 
```{r}
# (Only available on website)
# All interviewed households and group quarters. 
#
# Note the NIU on the codes page, this is a household variable and the
# NIU cases are the vacant households.
```

C) How many people received food stamps in 2011? 

D) What proportion of the population received food stamps in
2011? 

```{r}
# We will be working with the FOODSTMP variable a lot, so 
# let's turn it into a factor
cps_data <- cps_data %>%
  mutate(FOODSTMP_factor = as_factor(FOODSTMP))

cps_data %>% 
  group_by(FOODSTMP_factor) %>%
  summarize(n_foodstamp = sum(WTSUPP)) %>%
  mutate(pct_foodstmp = n_foodstamp / sum(n_foodstamp))

```


### Using household weights (HWTSUPP)
Suppose you were interested not in the number of people living in
homes that received food stamps, but in the number of households
that were food stamp participants. To get this statistic you would
need to use the household weight.

In order to use household weight, you should be careful to select
only one person from each household to represent that household's
characteristics. You will need to apply the household weight
(HWTSUPP). 

A) How many households received food stamps in 2011?

B) What proportion of households received food stamps in 2011?

```{r}
cps_data %>% 
  group_by(SERIAL) %>%
  filter(row_number() == 1) %>%
  group_by(FOODSTMP_factor) %>%
  summarize(n_foodstamp = sum(WTSUPP)) %>%
  mutate(pct_foodstmp = n_foodstamp / sum(n_foodstamp))

```

### Analyze the Sample – Part II Relationships in the Data ###
A) What is the universe for EMPSTAT in 2011?
```{r}
# Age 15+
```

B) What are the possible responses and codes for the selfreported
HEALTH variable? 
```{r}
print_labels(cps_data$HEALTH)
```

C) What percent of people with ‘poor’ self-reported health are at
work? 
```{r}
cps_data %>%
  filter(HEALTH == 5) %>%
  summarize(emp_pct = weighted.mean(EMPSTAT == 10, WTSUPP))
```

D) What percent of people with ‘very good’ self-reported health
are at work? 
```{r}
cps_data %>%
  filter(HEALTH == 2) %>%
  summarize(emp_n = weighted.mean(EMPSTAT == 10, WTSUPP))
```

E) In the EMPSTAT universe, what percent of people:

i. self-report ‘poor’ health and are at work? 

ii. self-report ‘very good’ health and are at work? 

```{r}
cps_data %>%
  filter(AGE >= 15) %>%
  mutate(HEALTH_factor = as_factor(HEALTH)) %>% 
  group_by(HEALTH_factor) %>%
  summarize(emp_n = weighted.mean(EMPSTAT == 10, WTSUPP))
```

### Analyze the Sample – Part III Relationships in the Data
A) What is the universe for AHRSWORK? 
```{r}
# Civilians age 15+, at work last week
```

B) What are the average hours of work for each self-reported health
category? 
```{r}
cps_data %>% 
  filter(AGE >= 15 & AHRSWORKT < 999) %>%
  mutate(HEALTH_factor = as_factor(HEALTH)) %>% 
  group_by(HEALTH_factor) %>%
  summarize(mean_hours_worked = weighted.mean(AHRSWORKT, WTSUPP))
  
```

