---
title: "Value Labels"
author: "Minnesota Population Center"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{value-labels}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Value labels in the ripums package
Integrated variables in IPUMS data often have value labels, which attach
text labels to the values taken by a variable (for example 1 = "Excellent").
These values can be incredibly useful during analysis. The `ripums` package
does import the labels, but they can be hard to spot at first:

```{r}
library(ripums)
ddi <- read_ipums_ddi(ripums_example("cps_00015.xml"))
cps <- read_ipums_micro(ddi, verbose = FALSE)

cps
```

The first clue that some of the variables are labelled is the `<dbl+lbl>`
that appear below `STATEFIP`, `ASECFLAG` and other variables. The tibble package 
prints the variable's type information below the variable name. You also can use
the `is.labelled()` function to check if a variable is labelled.

```{r}
is.labelled(cps$STATEFIP)
sapply(cps, is.labelled)
```

There are a few options to see what the actual labels are:
```{r}
# Printing the variable directly (or a subset)
head(cps$MONTH)

# Just get the labels
ipums_val_labels(cps$MONTH)

# or if you're working interactively you can use ipums_view
# ipums_view(ddi)
```

# Why use the labelled class instead of base R's factors?
The usual way to connect numeric data to labels in R is in `factor` variables. 
Though this data type is more native to R, and more widely supported by R code,
it was designed to for efficient calculations in linear models, not as a general 
purpose value labelling system and so is missing important features that the 
value labels provided by IPUMS require.

Factors only allow for integers to be mapped to a text label, and these
integers have to be a count starting at 1. This doesn't work for IPUMS 
data because often we use the numeric values to have meaning that would
be lost if converted to factors. The variable `AGE` uses the value to mean
the actual age, but does have value labels indicating what age 0 means and
the top codes.

```{r}
head(cps$AGE)

cps$AGE_FACTOR <- as_factor(cps$AGE)
head(cps$AGE_FACTOR)
```

It may seem like the new `AGE_FACTOR` variable is okay, but it can be very 
confusing!
```{r}
mean(cps$AGE)

# mean(cps$AGE_FACTOR) # error because data is a factor not numeric

mean(as.numeric(cps$AGE_FACTOR)) # A common mistake

# The "more" correct way, but NA because of the text labels
mean(as.numeric(as.character(cps$AGE_FACTOR)))
```

Because the factor variable has to assign values starting at 1, but the
AGE variable started at 0, most values were 1 higher than they should
have been. Not all values are 1 higher though, because not all values
exist in the data, so 85, 90, and 99 are 82, 83 and 84 respectively.

Factors also require that every value be labelled, which is not always
true in IPUMS data. In the AGE variable, the only values with labels
are 0, 90 and 99. For all other values, there is not additional label
information.

# Is the labelled class a panacea? (Hint: no)
Though the labelled does express all of the meaning provided by IPUMS value
labels into R, many R functions cannot use them or even actively remove them
from the data.

```{r}
ipums_val_labels(cps$HEALTH)

HEALTH2 <- ifelse(cps$HEALTH > 3, 3, cps$HEALTH)
ipums_val_labels(HEALTH2)
```

Therefore, your first task when importing an IPUMS dataset will usually be 
to convert the labelled values to other data structures. The bad news is 
that there's no good automatic way to do this; a lot depends on how you plan to use
the variables in your analysis and your preferences. 

The good news is that the ripums package provides several functions to 
make this process easier.

I think it is easiest to learn them by seeing them in action, but here is a list:
- `as_factor()` (really from the haven package)
- `zap_labels()`
- `lbl_na_if()`
- `lbl_collapse()`
- `lbl_relabel()`
- `lbl_add()`
- `lbl_add_vals()`

# Workflow example using ripums functions
## `as_factor()`
The HEALTH variable is structured just like a factor and so can be converted directly.
The `as_factor()` function is the easiest way to do so.
```{r}
ipums_val_labels(cps$HEALTH)
cps$HEALTH <- as_factor(cps$HEALTH)
```

The ASECFLAG and MONTH variable can also be converted directly (which were included by default by 
the IPUMS extract engine, but isn't that useful because this dataset only has respondents 
from ASEC in March)
```{r}
cps$ASECFLAG <- as_factor(cps$ASECFLAG)
cps$MONTH <- as_factor(cps$MONTH)
```

`as_factor` works on data.frames by converting every labelled variable to a factor. 
However, this can create confusing variables like AGE_FACTOR from above, so this
isn't the best thing to do right away.

## `zap_labels()`
I may decide that for my analysis, the AGE variable is most useful as the numeric values.
The `zap_labels()` function removes the labels.
```{r}
cps$AGE <- zap_labels(cps$AGE)
```
The topcodes (which are only available on the CPS website) indicate that the values
80 actually indicates 80-84 and 85 indicates 85+, so another option would be to
convert them to a factor with age ranges.

## `lbl_clean()`
This extract only contains data from a few states and I don't want the factor to have
a level for the unused ones. The `lbl_clean()` function keeps only labels for values
in the current dataset and returns a labelled variable which we can convert to a factor.
```{r}
ipums_val_labels(cps$STATEFIP)
cps$STATEFIP <- lbl_clean(cps$STATEFIP)

ipums_val_labels(cps$STATEFIP)
cps$STATEFIP <- as_factor(cps$STATEFIP)
```

## `lbl_na_if()`
The INCTOT variable has 2 labelled values that are not actually incomes:
99999998 indicates "Missing" and 99999999 indicate "Not in Universe".
On the CPS website, the Universe tab indicates that the Universe for 2016
is respondents age 15+. Let's say for my analysis, I can treat these values
as missing. 

The `lbl_na_if()` function takes the variable and a function that refers to
.val and .lbl (the values and labels respectively) and returns an indicator
of whether to set those values to NA and remove the label. You can also use the 
`~` notation from the purrr package to create succinct annonymous functions.

The .val and .lbl only refer to values that already have labels, they do not apply 
to unlabelled values. See the `lbl_add()` and `lbl_add_vals()` functions below for
working with unlabelled values.
```{r}
ipums_val_labels(cps$INCTOT)
# Caution: R defaults to printing large numbers like 99999999 in rounded 
# exponential format (1e+08) but that's not how they are actually stored

# All of these are equivalent
INCTOT1 <- lbl_na_if(cps$INCTOT, ~.val >= 99999990)
INCTOT2 <- lbl_na_if(cps$INCTOT, ~.lbl %in% c("Missing.", "N.I.U. (Not in Universe)."))
INCTOT3 <- lbl_na_if(cps$INCTOT, function(.val, .lbl) {
  is_missing <- .val == 99999998
  is_niu <- .lbl == "N.I.U. (Not in Universe)."
  return(is_missing | is_niu)
})
# Because only missing and NIU are labelled, and lbl_na_if ignores unlabelled
# values, this also works:
INCTOT4 <- lbl_na_if(cps$INCTOT, ~TRUE) 

# Change to a factor in the original cps data.frame
cps$INCTOT <- lbl_na_if(cps$INCTOT, ~.val >= 9999990)
cps$INCTOT <- as_factor(cps$INCTOT)
```

## `lbl_collapse()`
The EDUC variable provides an example of a common IPUMS practice of grouping
categories together by the starting digits. For example, the value 10 indicates 
"Grades 1, 2, 3, or 4", and 11 - "Grade 1", 12 - "Grade 2", etc. 

Let's say that I only care about those more general categories provided
by the first 2 digits. The `lbl_collapse()` function allows me to provide a function
that takes .val and .lbl and returns the value to assign it to. If that code is already
used, then the all of the values assigned to it will get that label, otherwise the label
of the smallest value is used. Just like with `lbl_na_if()`, the purrr-style compact 
syntax using `~` functions is supported.

```{r}
ipums_val_labels(cps$EDUC)
# %/% is integer division, which divides by the number but doesn't keep the remainder
cps$EDUC <- lbl_collapse(cps$EDUC, ~.val %/% 10)

ipums_val_labels(cps$EDUC)

cps$EDUC <- as_factor(cps$EDUC)
```

## `lbl_relabel()`
Sometimes you may wish to move the labels into new categories. For example, the categories 
in MIGRATE1 may not quite map what I want to use in my analysis.

The `lbl_relabel()` function provides a more flexible way to group existing labelled
values into new ones. It takes a two-sided formula, where the left-hand side is a 
label (defined with the `lbl()` function) and the right hand side is an expression
that can use .val and .lbl to evaluate to a logical indicating which values should
be assigned to this label.

```{r}
ipums_val_labels(cps$MIGRATE1)

cps$MIGRATE1 <- lbl_relabel(
  cps$MIGRATE1,
  lbl(0, "NIU / Missing") ~ .val %in% c(0, 2, 9),
  lbl(1, "Stayed in state") ~ .val %in% c(1, 3, 4),
  lbl(2, "Moved to new state") ~ .val == 5,
  lbl(3, "Moved abroad") ~ .val == 6
)
cps$MIGRATE1 <- as_factor(cps$MIGRATE1)
```

## `lbl_add()` and `lbl_add_vals()`
It's harder to come up with real world examples of when these functions would be useful, 
but for completion's sake, these functions allow you to create labels for values that
aren't already labelled.
```{r}
x <- haven::labelled(
  c(100, 200, 105, 990, 999, 230),
  c(`Unknown` = 990, NIU = 999)
)

lbl_add(x, lbl(100, "$100"), lbl(105, "$105"), lbl(200, "$200"), lbl(230, "$230"))
lbl_add_vals(x, ~paste0("$", .))
```

## Ready for analysis
And now, after converting all those labels to factors, I'm ready for analysis! If you
think of any other helper functions that would be useful for dealing with labels please
let us know by filing an issue on [github](https://github.com/mnpopcenter/ripums/issues).

# Other resources
The haven package vignette 'semantics' has some more details about the motivation and
implementation of the labelled class. You can view it by running the command:
```r
vignette("semantics", package = "haven")
```

If the ripums `lbl_*` functions don't match your workflow, there also is a package called
[labelled](http://larmarange.github.io/labelled/articles/intro_labelled.html) 
which provides other methods for manipulating value labels. It is not installed
by ripums, but is available on CRAN via the following command:
```r
install.packages("labelled")
```
